---
title: "ESM 206 LAB 5"
author: "Mauricio Collado"
date: "10/29/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Lab objectives
- Learn parse dates with lubridate
- Get counts with count()
- Uncount rows
- One and two sample t test
- Create a heatmap woth geom_tile()

```{r, include= FALSE}
# include is about showing the inputs or not

# Read in data and attach packages
library(tidyverse)
library(here)
library(janitor)

lobster_abundance <- read_csv(here::here("data","lobster_abundance.csv"), na="-99999") %>% 
  janitor::clean_names()

# NA in -99999
# variable names in capital letter
# date us column
# looks ok...BUT no unique observations. find e the mean based in the counts

# every single observation has its own row.

```

Use tidyr ::uncount function to convert our lobsters data to frequency format to caase format



```{r}
lobster_tidy <- lobster_abundance %>% 
  tidyr::uncount(lobster_count)


# populate all the missing values


```


### exploratory data visualization

ig i ignore variables like date, how the analysios change

```{r}

ggplot(lobster_tidy, aes(x=site, y=size_mm)) + #what makes the diffences
  geom_jitter(aes(color=site), 
              alpha=0.5, 
              width = 0.2)
#Histogram

ggplot(lobster_tidy, aes(x=size_mm))+
  geom_histogram(aes(fill=site))+
  facet_wrap(~site, scales = "free") #compara graphs without same units etc mislead the audience

#which is the distribution? they seem normality


#


ggplot(lobster_tidy, aes(sample=size_mm))+
  geom_qq()+
    facet_wrap(~site)  #normal ditribution? semms!. the degault is normal d but we can test others!\
#even i fht actual sample doesnt look normal, it is comfortable to say our mean is normal. no matter eh population strcuture, if wr take more samples and plots more means, i witll be normality d applies
#cONCEPTURALITY MAKES SENSE> Yes. for ...lobters

#mean is reasonable to compare? it seems smaller samples also are normal....mean is useful to vompare? NOT JUST S BIG NUMNBER



```

#### CONVERT DATE TO COLUMN TO CLASS DATE



WE LL THE LIBREDATE PACKAGET TO CNVERT DATA FORMALTE AND PARSE MONTH AND YEAR



```{r}

lobster_date <- lobster_tidy %>% 
  mutate(
    date_new=lubridate::mdy(date)
  )


# separate month and year?



```


Now lests parase year and month with month () and year()


```{r}
lobster_parse_date <- lobster_date %>% 
  mutate(
    obs_month = lubridate:: month(date_new),
    obs_year=lubridate::year(date_new)
    )# parse month and label it name
  

# class(lobster_parse_date$obs_month)



```

groups of lobters base in hierachacl groupins


many way to get ounts. we will see the most common and useful. they count groupt htins, doest th countsm, saves resultas and doest modigy tyour dataset


```{r}

# first lets count lobters by year and month

lobster_ym <- lobster_parse_date %>% 
  dplyr::count(obs_year, obs_month)


lobster_y <- lobster_parse_date %>% 
  dplyr::count(obs_year)



lobster_site <- lobster_parse_date %>% 
  dplyr::count(site) #having data in tidy format brings me freely to group


```


if we we stats over counts. it easier to usa group_by or n()


```{r}

lobster_summary <- lobster_parse_date %>% 
  group_by(site) %>% 
  summarize(
    mean_size=mean(site_mm, na.rm=TRUE),
    sd_size=sd(size, na.rm = TRUE),
    obster_number=n(),
    lobster_number=n()
  )



```

#### FIND CONFIDENCE INTERVALS


USE TTEST FUNCTION TO FINE CONDIDENC EINTERVALES AND PERFORM TE TESTS TO COMPARE MEAN IIN TOW SAMPLES.
```{r}


# if we calculta a 9%% confidence. it took a bunc more of the obervations. 


# we will use t student distribution. itd a normal distribution but with uncertainty into accoutn



ivee_lobster <- lobster_tidy %>% 
  filter(site=="IVEE") %>% 
  pull(size_mm)

# pull functions to extrat nd keep


t.test(ivee_lobster)


# the true mean of lengt of lobters is zeroooo? te score how different is the valye. p value tell us if ho is
# its mean is zero the probability that i can take a sample is nothing. Ther eisd chance that obters is zero ivee

# the chance so happening is paractical zero. 

# confiden interval means, more obs 95% will hace the true mean between 72 y 74 BUT htey miht be correct.

# here is a interval i would expect the 8% wlkk cobtai the true population mean! BUT IAM NOT SURe

# if 99.0 contains the ture value. we will have a wider view

```

THERE ARE SIGNIFICANT DIFFERENCE BETWEEN LOBSTER SIZE IN SITES. THE MEAN IS RELEVANTE?


```{r}

#2 SAMPLE T TEST. TH TI AVE ENOUGHT EVIDENCE THAT SAMPLES CAME FOROM POPULTATIONS WITH DIFFERENT MEANS.
# IF I DRAW TWO SAMPLES FROM THE SAMEE POPULATION AND ARE PERFECT RAMDON SAMPLES. THE DIFFERENCE WOULD BE ZEOOOOO
# WE DONE ENOIUGH EXPLORATORY ABNALYSI TO DETERMINA THAT A 2 SAMPLE T TEST FOR MEANS COMPARISON IS APPROPRIATE

# we ask their dirextion


napl_sample <- lobster_tidy %>% 
  filter(site=="NAPL") %>% 
  pull(size_mm)

mohk_sample <- lobster_tidy %>% 
    filter(site=="MOHK") %>% 
  pull(size_mm)

mn_ttest <- t.test(mohk_sample, napl_sample)
mn_ttest




# it IA THEDIRETION and otrder!!!!

# likely by random chance to pick observations from pop wih same mean. same mean and kind differet OR is really unlikely that they come for pops with same mean.

# its so unlikely to drawn the to sample, more likely that the started from different pop.





```

# the irs a significant differenc in lobster lengts betwee nbaples and mohawk reef


# this is the least interestin, more interesting, structkry of the data, mean values. 


# p value ois interes that there more interesting things ()t(df)=stats, p<0.05


```{r}


lobster_m <- lobster_tidy %>% 
  filter(site %in% c("NAPL", "MOHK"))

mn_ttest2 <- t.test(size_mm~ site, data=lobster_m)

mn_ttest2 
mn_test2$p.value
mn_test2$statistic
```

# there is a sig diff t(1850.9)=-19.548. p= 2.2.) we need to call parts


#?t.test


Here is the t-statistic : `r mn_tttest2$statistic`


```{r}

lobster_ys <- lobster_parse_date%>% 
  count(obs_year, site)


ggplot(data=lobster_ys, aes(x=obs_year, y=site))+
  geom_tile(aes(fill = n))


```

